```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F)

```

# Week 13 {-} 
<div style = "font-size: 28pt"> **_Multispecies models of competition: apparent competition_**</div>

## Lecture in a nutshell {-}

* **Model derivation**
<div style="margin-left: 30%;">$\begin{align}\frac {dN_{1}}{dt} = r_{1}N_{1}(1-\frac{N_{1}}{K_{1}})-a_{1}N_{1}P\end{align}\\$</div>
<div style="margin-left: 30%;">$\begin{align}\frac {dN_{2}}{dt} = r_{2}N_{2}(1-\frac{N_{2}}{K_{2}})-a_{2}N_{2}P\end{align}\\$</div>
<div style="margin-left: 30%; margin-bottom: 15px;">$\begin{align}\frac {dP}{dt} = e_{1}a_{1}N_{1}P+e_{2}a_{2}N_{2}P-mp\end{align}$</div>
    
    1. Find the equilibrium points:
        * $E_{1} = (\frac{d}{e_{1}a_{1}}, 0, \frac{r_{1}}{a_{1}}(1-\frac{d}{e_{1}a_{1}K_{1}}))$
        * $E_{2} = (0, \frac{d}{e_{2}a_{2}}, \frac{r_{2}}{a_{2}}(1-\frac{d}{e_{2}a_{2}K_{2}}))$
        * $E_{c} = (\frac{d-e_{2}a_{2}K_{2}(1-\frac{a_{2}r_{1}}{r_{2}a_{1}})}{e_{1}a_{1}+e_{2}a_{2}\frac{K_{2}a_{2}r_{1}}{K_{1}r_{2}a_{1}}}, \frac{d-e_{1}a_{1}K_{1}(1-\frac{a_{1}r_{2}}{r_{1}a_{2}})}{e_{2}a_{2}+e_{1}a_{2}\frac{K_{1}a_{1}r_{2}}{K_{2}r_{1}a_{2}}}, \frac{r_{1}}{a_{1}}(1-\frac{N_{1}^{*}}{K_{1}}))$ 
    2. Invasion analysis:
        * $N_{2}^{*} \rightarrow E_{1}$: $IGR_{2} = r_{2}-a_{2}[\frac{r_{1}}{a_{1}}(1-\frac{d}{e_{1}a_{1}K_{1}})] > 0$
        * $N_{1}^{*} \rightarrow E_{2}$: $IGR_{1} = r_{1}-a_{1}[\frac{r_{2}}{a_{2}}(1-\frac{d}{e_{2}a_{2}K_{2}})] > 0$
        * Mutual invasion criterion (coexistence): $\frac{r_{1}}{a_{1}}(1-\frac{d}{e_{1}a_{1}K_{1}})<\frac{r_{2}}{a_{2}}<\frac{r_{1}}{a_{1}}(1-\frac{d}{e_{2}a_{2}K_{2}})^{-1}$
    3. Local stability analysis:
        * $J_{E_{1}} = \begin{vmatrix} -\frac{r_{1}}{K_{1}}N_{1}^{*} & 0 & -a_{1}N_{1}^{*}\\0 & r_{2}-a_{2}P^{*} & 0\\e_{1}a_{1}P^{*} & e_{2}a_{2}P^{*} & 0\end{vmatrix}$ 
            1. Characteristic equation: $(r_{2}-a_{2}P^{*}-\lambda)[\lambda^{2}+\frac{r_{1}}{K_{1}}N_{1}^{*}\lambda+e_{1}a_{1}^{2}N_{1}^{*}P^{*}] = 0$
            2. $E_{1}$ is stable if $r_{2}-a_{2}P^{*}<0$ (i.e., $IGR_{2} <0$)
        * Same procedure for $E_{2}$
        * $J_{E_{c}} = \begin{vmatrix} -\frac{r_{1}}{K_{1}}N_{1}^{*} & 0 & -a_{1}N_{1}^{*}\\0 & -\frac{r_{2}}{K_{2}}N_{2}^{*} & -a_{2}N_{2}^{*}\\e_{1}a_{1}P^{*} & e_{2}a_{2}P^{*} & 0\end{vmatrix}$ 
            1. Characteristic equation: $\lambda^{3}+[\frac{r_{1}}{K_{1}}N_{1}^{*}+\frac{r_{2}}{K_{2}}N_{2}^{*}]\lambda^{2}+[\frac{r_{1}r_{2}}{K_{1}K_{2}}N_{1}^{*}N_{2}^{*}+e_{1}a_{1}^{2}N_{1}^{*}P^{*}+e_{2}a_{2}^{2}N_{2}^{*}P^{*}]\lambda\\+[e_{1}a_{1}^{2}\frac{r_{2}}{K_{2}}+e_{2}a_{2}^{2}\frac{r_{1}}{K_{1}}]N_{1}^{*}N_{2}^{*}P^{*} = 0$
            2. Using the Routh–Hurwitz stability criterion to determine the sign of the eigenvalues, we arrive at the conclusion that $E_{c}$ is locally stable if it is itself feasible.

<br>

## Lab demonstration {-}

Here, we are going to visualize the outcomes of apparent competition between two logistically-growing prey species _N~1~_ and _N~2~_ under linear predator consumption in the _r~1~/a~1~ - r~2~/a~2~_ space:

<div style="margin-left: 30%;">$\begin{align}\frac {dN_{1}}{dt} = r_{1}N_{1}(1-\frac{N_{1}}{K_{1}})-a_{1}N_{1}P\end{align}\\$</div>
<div style="margin-left: 30%;">$\begin{align}\frac {dN_{2}}{dt} = r_{2}N_{2}(1-\frac{N_{2}}{K_{2}})-a_{2}N_{2}P\end{align}\\$</div>
<div style="margin-left: 30%; margin-bottom: 15px;">$\begin{align}\frac {dP}{dt} = e_{1}a_{1}N_{1}P+e_{2}a_{2}N_{2}P-mp\end{align}$</div>

```{r, fig.width = 4, fig.height = 4, out.width = "75%", fig.align = "center", results = "hide", cache = TRUE}
library(tidyverse)
library(deSolve)

### 1. Create a function to evaluate the competition outcome
Apparent_logistic_func <- function(r1, r2){
  
  # specify the model 
  Apparent_logistic_model <- function(times, state, parms) {
    with(as.list(c(state, parms)), {
      dN1_dt = r1*N1*(1-N1/K1) - a1*N1*P
      dN2_dt = r2*N2*(1-N2/K2) - a2*N2*P
      dP_dt = e1*a1*N1*P + e2*a2*N2*P - m*P
      return(list(c(dN1_dt, dN2_dt, dP_dt)))  
    })
  }
  
  # model parameters
  times <- seq(0, 1000, by = 1)
  state <- c(N1 = 5, N2 = 5, P = 1)
  parms <- c(r1 = r1, r2 = r2, a1 = 0.35, a2 = 0.35, e1 = 0.4, e2 = 0.5, m = 0.8, K1 = 15, K2 = 12)
  
  # run the ode
  pop_size <- ode(func = Apparent_logistic_model, times = times, y = state, parms = parms)
  
  # get the final population size of N1 and N2
  N_final <- nrow(na.omit(pop_size))   # remove NAN's
  N1_final <- pop_size[N_final, 2]
  N2_final <- pop_size[N_final, 3]
  
  print(pop_size[N_final, ])
  
  # determine the outcome 
  if (N1_final < 1e-5) {
    return("N2_win")
  } else if (N2_final < 1e-5) {
    return("N1_win")
  } else {
    return("Coexist")
  }
  
}

### 2. Apply the function

# the parameter space to simulate over and the resolution
r_min <- 0
r_max <- 2.5
n <- 300

start_time <- Sys.time()  # the starting time of the simulations

comp_out <- expand.grid(r1 = seq(r_min, r_max, length.out = n),
                        r2 = seq(r_min, r_max, length.out = n)) %>%
  mutate(Outcome = map2_chr(r1, r2, Apparent_logistic_func))  # this is where the simulations actually take place!

end_time <- Sys.time()  # the ending time of the simulations

```

```{r}
end_time - start_time  # it might take a while!
```

```{r, fig.width = 4, fig.height = 4, out.width = "75%", fig.align = "center"}
library(ggplot2)

### 3. Visualize the competition outcomes in the grids
ggplot(comp_out, aes(xmin = r1/0.5, ymin = r2/0.5, 
                     xmax = r1/0.5 + (r_max-r_min)/(n-1)/0.5, ymax = r2/0.5 + (r_max-r_min)/(n-1)/0.5,
                     fill = Outcome)) + 
  geom_rect(show.legend = F, alpha = 0.9) + 
  geom_abline(intercept = 0, slope = (1-0.8/(0.4*0.35*15)), size = 1.2) + 
  geom_abline(intercept = 0, slope = (1-0.8/(0.5*0.35*12))^-1, size = 1.2) + 
  labs(x = expression(r[1]/a[1]), y = expression(r[2]/a[2])) + 
  scale_x_continuous(limits = c(r_min/0.5, (r_max+(r_max-r_min)/(n-1))/0.5 ), expand = c(0, 0)) +
  scale_y_continuous(limits = c(r_min/0.5, (r_max+(r_max-r_min)/(n-1))/0.5 ), expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set1") + 
  theme_bw(base_size = 14) + 
  theme(panel.grid = element_blank()) +
  coord_fixed(ratio = 1) + 
  annotate(geom = "text", x = 1, y = 4, label = "bold(bolditalic(N[2])~wins)", parse = T) + 
  annotate(geom = "text", x = 4, y = 4, label = "bold(Coexist)", parse = T) + annotate(geom = "text", x = 4, y = 1, label = "bold(bolditalic(N[1])~wins)", parse = T)

```

Note that the area of coexistence depends on the slopes of the boundaries, which are $1-\frac{m_{1}}{e_{1}a_{1}K_{1}}$ between **_N~1~ wins_** and **_Coexist_**, and $(1-\frac{m_{2}}{e_{2}a_{2}K_{2}})^{-1}$ between **_N~2~ wins_** and **_Coexist_**.

<br>

## Additional readings {-}

[Predation, apparent competition, and the structure of prey communities](./Additional readings/Holt_1977_Theo_Pop_Bio.pdf){target="_blank"}
<br>

## Assignments {-}

No assignments this week.


