```{r setup, include = FALSE}
knitr::opts_chunk$set(message = F, 
                      error = F, 
                      warning = F)

```

# Week 14 {-} 
<div style = "font-size: 28pt"> **_Disease dynamics and SIR models_**</div>

## Lecture in a nutshell {-}

* **Disease dynamics**
    1. Compartmental models: assign individuals to different categories and model the transitions between them
    2. Density-based transmission vs. Frequency-based transmission: whether the rate of contacting other individuals scales with the population density. For example, in the SI model without demography,
        * Density-based transmission: the growth of the infected individuals is $\beta SI$
        * Frequency-based transmission: the growth of the infected individuals is $\frac{\beta SI}{N}$

<div style="height:1px ;"><br></div>

* **SI model without demography**
    1. System of equations:
<div style="margin-left: 35%;">$\begin{align}\frac {dS}{dt} =-\beta SI\end{align}\\$</div>
<div style="margin-left: 35%;">$\begin{align}\frac {dI}{dt} = \beta SI\end{align}\\$</div>
    2. The number of infected individuals will grow logistically and the disease will eventually infect the entire population. 
    
<div style="height:1px ;"><br></div>    

* **SIR model without demography**
    1. System of equations:
<div style="margin-left: 35%;">$\begin{align}\frac {dS}{dt} =-\beta SI\end{align}\\$</div>
<div style="margin-left: 35%;">$\begin{align}\frac {dI}{dt} = \beta SI-\rho I\end{align}\\$</div>
<div style="margin-left: 35%; margin-bottom: 15px;">$\begin{align}\frac {dR}{dt} = \rho I\end{align}$</div>
    2. $R_{0}$: the number of secondary infections caused by one infected individual in a fully susceptible population (= per capita infection in a fully susceptible population*infectious time)
    3. For SIR model without demography, $R_{0}$ is $\frac{\beta}{\rho}S_{0}$
    4. Can the disease spread at $S_{0}$? The invasion growth rate of _I_ is $\lim_{I \to 0} \frac{dI}{dt}\frac{1}{I} = \beta S_{0}-\rho$
        * If $R_{0}$ > 1, IGR > 0, the disease will spread initially but later decline and eventually self-extinguish. Part of the population will be free from the infection. _S_ and _R_ coexist at the equilibrium.
        * If $R_{0}$ < 1, IGR < 0, disease cannot spread and will just gradually die out. The population will consist of mostly _S_ and a few _R_ at the equilibrium.

<div style="height:1px ;"><br></div>    

* **SIR model with demography**
    1. System of equations:
<div style="margin-left: 35%;">$\begin{align}\frac {dS}{dt} = \theta-\beta SI-\delta S\end{align}\\$</div>
<div style="margin-left: 35%;">$\begin{align}\frac {dI}{dt} = \beta SI-\rho I-\gamma I\end{align}\\$</div>
<div style="margin-left: 35%; margin-bottom: 15px;">$\begin{align}\frac {dR}{dt} = \rho I-\delta R\end{align}$</div>
    2. $R_{0}$ is $\frac {\beta}{\rho+\gamma}\frac{\theta}{\delta}$
    3. The equilibrium points $(S^{*}, I^{*}, R^{*})$:
        * $E_{DF}$ (disease-free equilibrium) = $(\frac{\theta}{\delta}, 0, 0)$
        * $E_{E}$ (endemic equilibrium) = $(\frac{\gamma + \rho}{\beta}, \frac{\theta \beta-\delta(\gamma+\rho)}{\beta(\gamma+\rho)}, (\frac {\beta}{\rho+\gamma}\frac{\theta}{\delta}-1)\frac{\rho}{\beta})$ = $(\frac{1}{R_{0}}\frac{\theta}{\delta}, (R_{0}-1)\frac{\delta}{\beta}, (R_{0}-1)\frac{\rho}{\beta})$
    4. Invasion analysis: $\lim_{I \to 0} \frac{dI}{dt}\frac{1}{I} = \beta S_{DF}-\rho - \gamma$; disease can spread if $\beta \frac{\theta}{\delta} -\rho - \gamma > 0$ (or $R_{0} > 1$)
    5. Local stability analysis:
        * $E_{DF}$
            1. $J_{DF} = \begin{vmatrix} -\delta & -\beta S^{*} & 0\\0 & \beta S^{*}-\rho-\gamma & 0\\0 & \rho & -\delta\end{vmatrix}$ 
            2. Eigenvalues: $-\delta$, $-\delta$, $\beta S^{*}-\rho-\gamma$
            3. Locally stable if $\beta S^{*}-\rho-\gamma < 0$, that is, $R_{0} < 1$
        
        * $E_{E}$
            1. $J_{E} = \begin{vmatrix} -\beta I^{*}-\delta & -\beta S^{*} & 0\\\beta I^{*} & 0 & 0\\0 & \rho & -\delta\end{vmatrix}$  
            2. Characteristic equation: $(\delta + \lambda)(\lambda^{2}+(\beta I^{*}+\delta)\lambda+\beta^{2}I^{*}S^{*}) = 0$
            3. Locally stable if $I^{*} > 0$ ($I^{*}$ is feasible), that is, $R_{0} > 1$

<br>

## Lab demonstration {-}

In today's lab, we are going to simulate the SIR model with demography and visualize two types of disease dynamics: (1) the endemic equilibrium, at which the S, I, and R all coexist, and (2) the disease-free equilibrium, at which the disease will die off and only S remains.

<div style="margin-left: 35%;">$\begin{align}\frac {dS}{dt} = \theta-\beta SI-\delta S\end{align}\\$</div>
<div style="margin-left: 35%;">$\begin{align}\frac {dI}{dt} = \beta SI-\rho I-\gamma I\end{align}\\$</div>
<div style="margin-left: 35%; margin-bottom: 15px;">$\begin{align}\frac {dR}{dt} = \rho I-\delta R\end{align}$</div>

```{r, fig.width = 5, fig.height = 3.5, out.width = "70%", fig.align = "center"}
library(tidyverse)
library(deSolve)

SIR_model_fun <- function(theta, beta, delta, rho, gamma, title){
  
  # model specification
  SIR_model <- function(times, state, parms) {
    with(as.list(c(state, parms)), {
      dS_dt = theta - beta*S*I - delta*S
      dI_dt = beta*S*I - rho*I - gamma*I 
      dR_dt = rho*I - delta*R
      return(list(c(dS_dt, dI_dt, dR_dt)))
    })
  }
  
  # model parameters
  times <- seq(0, 1000, by = 1)
  state <- c(S = 25, I = 0.1, R = 0)
  parms <- c(theta = theta, beta = beta, delta = delta, rho = rho, gamma = gamma)
  
  # model application
  SIR_out <- ode(func = SIR_model, times = times, y = state, parms = parms)
  
  # visualization
  SIR_out %>%
    as.data.frame() %>%
    pivot_longer(cols = -time, names_to = "state", values_to = "N") %>%
    mutate(state = fct_relevel(state, "S", "I", "R")) %>%
    ggplot(aes(x = time, y = N, color = state)) + 
    geom_line(size = 1.5) +
    theme_classic(base_size = 14) +
    labs(x = "Time", y = "N", title = title) +
    scale_x_continuous(limits = c(0, 1010), expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 55), expand = c(0, 0)) +
    scale_color_brewer(name = NULL, palette = "Set1") + 
    theme(plot.title = element_text(hjust = 0.5))
  
}
```

```{r, fig.width = 5, fig.height = 3.5, out.width = "70%", fig.align = "center"}
### Epidemic equilibrium  
SIR_model_fun(theta = 0.25, beta = 0.01, delta = 0.01, rho = 0.02, gamma = 0.02, title = "Epidemic equilibrium")

### Disease-free equilibrium
SIR_model_fun(theta = 0.25, beta = 0.01, delta = 0.01, rho = 0.3, gamma = 0.02, title = "Disease-free equilibrium")

```

By increasing the recovery rate $\rho$ in the second example, we drive the basic reproduction number $R_{0}$ below 1, and thus the disease will not spread and the system reaches the disease-free equilibrium.

<br>

## Additional readings {-}

[Population biology of infectious diseases: Part I](./Additional readings/Anderson_&_May_1979_I_Nature.pdf){target="_blank"}

[Population biology of infectious diseases: Part II](./Additional readings/Anderson_&_May_1979_II_Nature.pdf){target="_blank"}

<br>

## Assignments {-}

[Graphical and Local Stability Analysis of the SIR Model](./Assignments/Week14_SIR Model.pdf){target="_blank"}

[Suggested Solutions](./Assignments/Week14_SIR Model_with_Solutions.pdf){target="_blank"}





